### 2025-03-11 paper/preprint/paper.tex
Add first part of discussion (Section 5.1)

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\subsubsection{Impact of hyperparameter
settings}\label{impact-of-hyperparameter-settings}

@@

\subsection{Approach/Future Directions}\label{approachfuture-directions}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Limited to classification models.
\item
  Training instabilities.
\item
  Hyperparameter sensitivity -\textgreater{} can we do better than grid
  search? (Bayes opt, \ldots)
\end{enumerate}

\subsection{Limitations}\label{limitations}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Proxy attributes of immutable features.
\item
  Increased training time.
\item
  Fairness and caveats (aware it's not a classical approach in this
  context, but there is a clear link).
\end{enumerate}

@@

@@

@@

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\subsubsection{Impact of hyperparameter
settings}\label{sec-hyperparameters}

@@

We begin the discussion by addressing the direct extensions of the
counterfactual training approach in Section~\ref{sec-approach}. Then, we
look at its broader limitations and challenges in
Section~\ref{sec-limitations}.

\subsection{Future research}\label{sec-approach}

\textbf{\emph{CT is defined only for classification settings.}} Our
formulation relies on the distinction between non-target class(es)
\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
arbitrarily defined by the user, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it applies to binary and
multi-class classification but it is not well-defined for other ML tasks
where the change in outcome with respect to a decision threshold
\(\tau\) cannot be readily quantified. In fact, this is a common
restriction in research on CEs and AR that predominantly focuses on
classification models. Although other settings have attracted some
interest (e.g., regression in
\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
\citeproc{ref-zhao2023counterfactual}{Zhao, Broelemann, and Kasneci
2023}), there is still no consensus on what constitutes a counterfactual
in such settings.

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter major
variability in the outcomes. CT is exposed to two potential sources of
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
For example, Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023})
recognize this to be a challenge for \emph{ECCCo} and so it may have
downstream impacts on our proposed method. Still, we find that training
instabilities can be largely counteracted with strong regularization, in
particular through \(\lambda_{\text{reg}}\) penalty on the energy.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
discussed in Section~\ref{sec-hyperparameters}, our method requires
careful selection of the hyperparameters. All experiments relied on
grid-search tuning but we recognize that other algorithms may be more
efficient. Notably, counterfactual training is iterative which makes a
variety of methods applicable, including Bayesian (e.g.,
\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
al. 2017}) optimization.

\subsection{Limitations and challenges}\label{sec-limitations}

\textbf{\emph{CT increases the training time of models.}}

\textbf{\emph{Immutable features may have proxies.}}

\textbf{\emph{Interventions on features may have downstream impacts on
fairness.}}

@@

\bibitem[\citeproctext]{ref-franceschi2017forward}
Franceschi, Luca, Michele Donini, Paolo Frasconi, and Massimiliano
Pontil. 2017. {``{Forward and Reverse Gradient-Based Hyperparameter
Optimization}.''} In \emph{Proceedings of the 34th International
Conference on Machine Learning}, edited by Doina Precup and Yee Whye
Teh, 70:1165--73. Proceedings of Machine Learning Research. PMLR.
\url{https://proceedings.mlr.press/v70/franceschi17a.html}.

@@

\bibitem[\citeproctext]{ref-snoek2012practical}
Snoek, Jasper, Hugo Larochelle, and Ryan P. Adams. 2012. {``{Practical
Bayesian Optimization of Machine Learning Algorithms}.''} In
\emph{Advances in Neural Information Processing Systems}, edited by F.
Pereira, C. J. Burges, L. Bottou, and K. Q. Weinberger. Vol. 25. Curran
Associates, Inc.
\url{https://proceedings.neurips.cc/paper_files/paper/2012/file/05311655a15b75fab86956663e1819cd-Paper.pdf}.

\bibitem[\citeproctext]{ref-spooner2021counterfactual}
Spooner, Thomas, Danial Dervovic, Jason Long, Jon Shepard, Jiahao Chen,
and Daniele Magazzeni. 2021. {``{Counterfactual Explanations for
Arbitrary Regression Models}.''} \emph{CoRR} abs/2106.15212.
\url{https://arxiv.org/abs/2106.15212}.

@@

\bibitem[\citeproctext]{ref-zhao2023counterfactual}
Zhao, Xuan, Klaus Broelemann, and Gjergji Kasneci. 2023.
{``{Counterfactual Explanation for Regression via Disentanglement in
Latent Space}.''} In \emph{2023 IEEE International Conference on Data
Mining Workshops (ICDMW)}, 976--84. Los Alamitos, CA, USA: IEEE Computer
Society. \url{https://doi.org/10.1109/ICDMW60847.2023.00130}.

### 2025-03-11 paper/sections/discussion.qmd
latex is going to kill me one day
phew
vamos
some notes from meeting and updates to appendix
Add first part of discussion (Section 5.1)

1. Limited to classification models.
2. Proxy attributes of immutable features.

>>>>

We begin the discussion by addressing the direct extensions of the counterfactual training approach in @sec-approach. Then, we look at its broader limitations and challenges in @sec-limitations.

## Future research {#sec-approach}

***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined by the user, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it applies to binary and multi-class classification but it is not well-defined for other ML tasks where the change in outcome with respect to a decision threshold $\tau$ cannot be readily quantified. In fact, this is a common restriction in research on CEs and AR that predominantly focuses on classification models. Although other settings have attracted some interest [e.g., regression in @spooner2021counterfactual; @zhao2023counterfactual], there is still no consensus on what constitutes a counterfactual in such settings.

***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method. Still, we find that training instabilities can be largely counteracted with strong regularization, in particular through $\lambda_{\text{reg}}$ penalty on the energy.

***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method requires careful selection of the hyperparameters. All experiments relied on grid-search tuning but we recognize that other algorithms may be more efficient. Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

## Limitations and challenges {#sec-limitations}

***CT increases the training time of models.***

***Immutable features may have proxies.***

***Interventions on features may have downstream impacts on fairness.***
